---
title: Introduction
execute: 
  eval: false
---

## Background

Denitrification is the natural process by which nitrate (NO~3~^-^) in the soil are converted by bacteria into nitrous oxide (N~2~O) or pure nitrigen (N~2~). The latter is called *total denitrification* --- the full process described in @eq-denitrification takes place.

$$
\ce{NO3^- ->[\text{Nitrate}][\text{reductase}] NO2^- ->[\text{Nitrite}][\text{reductase}] NO ->[\text{Nitrite oxide}][\text{reductase}] N2O^- ->[\text{Nitrous oxide}][\text{reductase}] N2}
$${#eq-denitrification}

Denitrification occurs in conditions where oxygen is limited, such as waterlogged soils. It is part of the nitrogen cycle, where nitrogen is circulated between the atmosphere, organisms and the earth.

## Installation of the R package PRE

All functions that are used repeatedly throughout this projec are implemented in an R package called `PRE`. The source code of this package is stored in a [publically accessible way on GitHub](https://github.com/Damian-Oswald/PRE). You can install the package directly from GitHub using the `install_github` function from the `remotes` package.

```{r}
install.packages("remotes")
remotes::install_github("https://github.com/Damian-Oswald/PRE")
```

Alternatively, you could download the GitHub repository and manually install the `PRE` package locally.
Once the package is installed, you can attach it to the search path in the usual way.

```{r}
#| eval: true
#| message: false
library(PRE)
```

### What do the various functions do?

The R code shown on this project help page is extremely simplified. If you want want to understand what a specific function of the `PRE` package does, simple visit the corresponding R help page. For example, you could read on the function `crossValidate` by using one of the the following commands.

```{r}
help("crossValidate", package = "PRE")
?crossValidate
```

If you want to jump really deep into a function, you could print it out in your console.

```{r}
#| eval: true
print(crossValidate)
```

Note this function source code output is lacking any comments. Those were automatically removed when compiling the R package. To see the commented source code, you'll have to jump into the package files, or visit the GitHub page. Simply use the following URL:

<https://github.com/Damian-Oswald/PRE/blob/main/R/crossValidate.R>

Notice how the base name (the last part of the URL) matches the function name. You can replace that by any function name of your liking.
