# Results

## Summary table

```{r}
#| echo: false
#| tbl-colwidths: [22,14,16,16,16,16]
#| label: tbl-by-depth-and-variety
#| tbl-cap: "Overall results of the process rate estimator (μ ±  σ) by depth layer and variety. N~2~O~nit~ is the sum of the nitrification rate over the entire estimated period, while N~2~O~den~ is the sum of the denitrification rate. For reference, this table also displays N~2~O~produced~, which is simply N~2~O~nit~ + N~2~O~den~, as well as N~2~O~consumed~, which is the sum of the reduction rate."

# read the results
data <- read.csv(file.path("scripts","run-process-rate-estimator","output","estimated-process-rates.csv"))

# compute the sum of process rates over entire time
for (x in c("Nitrification", "Denitrification", "Reduction")) {
    assign(x, 0.001 * tapply(data[,x], with(data, list(variety, depth, column)), PRE::nonNegative, f = sum))
}
names(dimnames(Nitrification)) <- names(dimnames(Denitrification)) <- names(dimnames(Reduction)) <- c("Variety", "Depth", "Column")

# compute total produced nitrogen
Production <- Nitrification + Denitrification

# # STATISTICAL TESTS:
# # ==================
# for (j in c("Nitrification", "Denitrification", "Production", "Reduction")) {
#     df <- na.omit(reshape2::melt(get(j), value.name = "Process"))
#     for (i in c("Variety", "Depth")) df[,i] <- as.factor(df[,i])
#     model <- lm(Process ~ Depth, data = df)
#     cat(paste0(j,":\n",strrep("=",nchar(j)+1),"\n"))
#     print(kruskal.test(Process ~ Variety, data = df))
#     print(kruskal.test(Process ~ Depth, data = df))
# }

# write function to paste mean ± sd
f <- function(x) paste(signif(mean(x, na.rm = TRUE), 2), "±", signif(sd(x, na.rm = TRUE), 2))

# bind results to data frame
df <- cbind(reshape2::melt(apply(Nitrification, 1:2, f)),
            pod = reshape2::melt(apply(Denitrification, 1:2, f))[,3],
            red = reshape2::melt(apply(Production, 1:2, f))[,3],
            den = reshape2::melt(apply(Reduction, 1:2, f))[,3])
df <- df[order(df[,"Variety"]),]
colnames(df) <- c("Variety", "Depth [cm]", "N~2~O~nit~", "N~2~O~den~", "N~2~O~produced~", "N~2~O~consumed~")
df[,1] <- as.character(df[,1])
df[((1:nrow(df))[-c((0:4)*5+1)]),1] <- ""
knitr::kable(df, row.names = FALSE, align = "llrrrr")
```

Owing to all processes' significant divergence of the linear model's residuals from a normal distribution, as determined by the Shapiro-Wilk normality test, the Kruskal-Wallis Rank Sum Test was employed to examine group variation among the varieties. These respective tests were performed utilizing the relevant functions -- `shapiro.test` and `kruskal.test`, from the `stats` package, as referred to in [@r2023language].

Among the different varieties, there were no significant group differences detected for neither N~2~O~nit~ nor N~2~O~den~, and hence also not for N~2~O~produced~. However, significant differences among the varieties for N~2~O~consumed~ were detected (p = 0.01098).

```{r}
#| echo: false
#| tbl-colwidths: [36,16,16,16,16]
#| label: tbl-by-variety
#| tbl-cap: "Values in @tbl-by-depth-and-variety aggregated by variety."

# bind results to data frame
df <- cbind(nit = reshape2::melt(apply(Nitrification, 1, f)),
            pod = reshape2::melt(apply(Denitrification, 1, f)),
            red = reshape2::melt(apply(Production, 1, f)),
            den = reshape2::melt(apply(Reduction, 1, f)))
df <- data.frame(rownames(df), df)
colnames(df) <- c("Variety", "N~2~O~nit~", "N~2~O~den~", "N~2~O~produced~", "N~2~O~consumed~")
knitr::kable(df, row.names = FALSE, align = "lrrrr")
```

Conversely, N~2~O~consumed~ did not show any significant group differences with respect to the depth layers. However, all other processes did: N~2~O~nit~ (p = 1.005e-07) nor N~2~O~den~ (p = 0.001029), and hence also for N~2~O~produced~ (p = 3.883e-08).

```{r}
#| echo: false
#| tbl-colwidths: [22,14,16,16,16,16]
#| label: tbl-by-depth
#| tbl-cap: "Values in @tbl-by-depth-and-variety aggregated by depth layer."

# bind results to data frame
df <- cbind(nit = reshape2::melt(apply(Nitrification, 2, f)),
            pod = reshape2::melt(apply(Denitrification, 2, f)),
            red = reshape2::melt(apply(Production, 2, f)),
            den = reshape2::melt(apply(Reduction, 2, f)))
df <- data.frame(x = rep("",5), d = rownames(df), df)
colnames(df) <- c("", "Depth [cm]", "N~2~O~nit~", "N~2~O~den~", "N~2~O~produced~", "N~2~O~consumed~")
knitr::kable(df, row.names = FALSE, align = "llrrrr")
```

## Processes over time

The tabset panels below show all results, from column 1 to 12[^missing] and depths 7.5 cm, 30 cm, 60 cm, 90 cm and 120 cm.

[^missing]: Note that due to missing measurements, column 6 couldn't be computed completely.

```{r}
#| eval: true
#| echo: false
#| results: asis
cat("::: {.panel-tabset}\n\n")
for (column in 1:12) {
    
    if(column==1) {
        cat("## Column 1\n\n")
    } else {
        cat("##", column, "\n\n")
    }
    
    cat("::: {.panel-tabset}\n\n")
    
    for (depth in PRE::getParameters()$depths) {
        
        if(depth==7.5) {
            cat("## Depth =", depth, "cm\n\n")
        } else {
            cat("##", depth, "cm\n\n")
        }
        
        cat(sprintf("![](scripts/run-process-rate-estimator/output/visualized-process-rates-C%s-D%s.svg){width=\"100%%\"}\n\n", column, depth))
    }
    cat(":::\n\n")
}

cat(":::\n\n")
```

## Analysis of variance

### Nitrification

```{r}
#| echo: false
#| results: asis
data <- read.csv(file.path("scripts","run-process-rate-estimator","output","estimated-process-rates.csv"))
df <- reshape2::melt(with(data, tapply(Nitrification, list(depth, column), mean)), varnames = c("depth", "column"), value.name = "Nitrification")
model <- lm(Nitrification ~ depth * as.factor(column), df)
options(knitr.kable.NA = "")
df <- as.data.frame(anova(model))
df <- data.frame(signif(df, 3), stars = gtools::stars.pval(df$`Pr(>F)`))
colnames(df)[5:6] <- c("p-value", " ")
knitr::kable(df, format = "pipe")
```

### Denitrification

```{r}
#| echo: false
#| results: asis
df <- reshape2::melt(with(data, tapply(Denitrification, list(depth, column), mean)), varnames = c("depth", "column"), value.name = "Denitrification")
model <- lm(Denitrification ~ depth, df)
df <- as.data.frame(anova(model))
df <- data.frame(signif(df, 3), stars = gtools::stars.pval(df$`Pr(>F)`))
colnames(df)[5:6] <- c("p-value", " ")
knitr::kable(df, format = "pipe")
```

### Reduction

```{r}
#| echo: false
#| results: asis
df <- reshape2::melt(with(data, tapply(Reduction, list(depth, column), mean)), varnames = c("depth", "column"), value.name = "Reduction")
model <- lm(Reduction ~ depth * as.factor(column), df)
df <- as.data.frame(anova(model))
df <- data.frame(signif(df, 3), stars = gtools::stars.pval(df$`Pr(>F)`))
colnames(df)[5:6] <- c("p-value", " ")
knitr::kable(df, format = "pipe")
```
