---
title: Running the process rate estimator
execute: 
  eval: false
  echo: true
---

## Install the R package `PRE`

You can directly install this R package from GitHub. The package itself already contains all data and parameters necessary to run the process rate estimator.

```{r}
install.packages("remotes")
remotes::install_github("https://github.com/Damian-Oswald/PRE")
```

Once installed, add the package to the search path.

```{r}
library(PRE)
```

## Prepare the data

In a first step, we can calculate the N~2~O-N from the original measurements.

```{r}
original <- getN2ON(data = measurements, parameters = parameters)
```

Next, the data are interpolated. This process uses kernel regression to interpolate between the measured values of N~2~O-N, SP and d18O.

```{r}
data <- getMissing(data = data, hyperparameters = hyperparameters)
```

With all of these steps done, we can calculate the Fluxes according to equations 1, 2, 3.

```{r}
data <- calculateFluxes(data = data, parameters = parameters)
```

Now, everything is ready to run the PRE.

## Running the process rate estimator


```{r}
# run the solver once
x <- PRE(data = data, column = 1, depth = 7.5, date = "2016-01-01", n = 200)

# print out information
print(x)

# plot
plot(x)

# pairs panel
pairs(x)
```