---
execute: 
  eval: false
---

# Sensitivity Analysis

Sensitivity analysis is a method used to determine how different values of an independent variable impact a particular dependent variable under a given set of assumptions. This technique is used in various fields such as finance, engineering, and environmental studies to test the robustness of models against changes in input values. By systematically varying key parameters, sensitivity analysis helps in identifying which variables have the most significant impact on the outcome, aiding in decision-making and risk assessment.

:::{.column-page}
| End Member             |                                 | Value [‰] | Value reference         | Sampling                         | Range reference   |
|------------------------|---------------------------------|-----------|-------------------------|----------------------------------|-------------------|
| `eta_SP_diffusion`     | $\eta\: \text{SP}_\text{diff.}$ | 1.55      | @well2008isotope        | $\sim \mathcal{N}(1.55, 0.28^2)$ | @well2008isotope  |
| `eta_18O_diffusion`    | $\eta\ce{^{18}O}_\text{diff.}$  | -7.79     | @well2008isotope        | $\sim \mathcal{N}(-7.79, 0.27^2)$| @well2008isotope  |
| `SP_nitrification`     | $\text{SP}_\text{nit.}$         | 34.4      | @decock2013potential    | $\sim \mathcal{U}(26.2, 34.6)$   | @denk2017nitrogen |
| `d18O_nitrification`   | $\delta\ce{^{18}O_\text{nit.}}$ | 36.5      | @lewicka2017quantifying | $\sim \mathcal{N}(36.5, 2^2)$    |                   |
| `SP_denitrification`   | $\text{SP}_\text{den.}$         | -2.4      | @decock2013potential    | $\sim \mathcal{U}(-2.4, -0.9)$   | @denk2017nitrogen |
| `d18O_denitrification` | $\delta\ce{^{18}O_\text{den.}}$ | 11.1      | @lewicka2017quantifying | $\sim \mathcal{N}(11.1, 2^2)$    |                   |
| `eta_SP_reduction`     | $\eta\:\text{SP}_\text{red.}$   | -5.3      | @denk2017nitrogen       | $\sim \mathcal{U}(-8, -2)$       | @denk2017nitrogen |
| `eta_18O_reduction`    | $\eta\ce{^{18}O_\text{red.}}$   | -16.1     | @lewicka2017quantifying |                                  |                   |

: Isotope end members values used for modeling gross nitrification derived N~2~O production, gross denitrification derived N~2~O production, and gross N~2~O reduction. @lewicka2017quantifying originally report $\delta \ce{^{18} O \text - N2O} (\ce{N2O}/\ce{H2O})$. When a range is indicated, a random number from the range was selected in each iteration. When an average and standard deviation is indicated, a random value was drawn from the normal distribution in each iteration. {#tbl-sensitivity tbl-colwidths="[15,10,10,20,15,20]"}
:::

To conduct the sensitivity analysis, a data frame with parameters sampled according to the distributions in @tbl-sensitivity is created. We'll have to specify how many times we want to do this --- in this example that's `n = 200`.

```{r}
n <- 200
parameters <- data.frame(eta_SP_diffusion = rnorm(n, 1.55, 0.28),
                         eta_18O_diffusion = rnorm(n, -7.79, 0.27),
                         SP_nitrification = runif(n, 26.2, 34.6),
                         d18O_nitrification = rnorm(n, 36.5, 2),
                         SP_denitrification = runif(n, -2.4, -0.9),
                         d18O_denitrification = rnorm(n,11.1, 2),
                         eta_SP_reduction = runif(n,-8,-2),
                         eta_18O_reduction = runif(n,-24,-6))
```

Next, we write a function `f` that will run the process rate estimator for a specified column and depth with the parameters in the vector `p` --- which coincides to one row in the `parameters` data frame.

```{r}
f <- function(p, data, column = 1, depth = 7.5) {
    x <- longPRE(data = data,
                 column = column,
                 depth = depth,
                 n = 3,
                 parameters = do.call(getParameters, as.list(p)))
    return(x[["processes"]])
}
```

That function makes it very straightforward to compute the data for the sensitivity analysis. We'll simply apply `f` to every single row in `parameters` using the `apply` function [@wickham2011split].

```{r}
results <- t(apply(parameters, 1, f, data = data, column = 1, depth = 7.5))
```

To conduct the sensitivity analysis, we now model the results based on the sampled parameters -- so we treat the sampled parameters as independent variables $\mathbf X$, and the computed process rates as dependent variables $\mathbf Y$.

First, we might look at the pairwise correlations between columns of $\mathbf X$ and $\mathbf Y$, i.e. all combinations of process rates as well as isotope end members (@fig-sensitivity).

:::{.column-screen-inset}
![Pairwise results of the process rates (y-axis) and the isotope end members (x-axis) for column 1 at 7.5 cm depth. The red lines indicate the respective linear models with their 95% conficence intervals.](graphics/sensitivity.svg){#fig-sensitivity width=100%}
:::

@fig-sensitivity already reveals quite a lot, but it would be better to construct a multiple linear regression model where we include all isotope end members at the same time.

The table below shows the resulting coefficients of such a multiple linear regression model. We can directly interpret the regression coefficients without any further modifications, as all considered variables share the same unit (‰).

```{r}
#| echo: false
#| eval: true
#| results: asis
#| tbl-cap: Regression results of multiple linear models with stepwise variable selection.
cat(readLines("resources/tbl-sensitivity.txt"))
```

Note that the coefficients of column, depth and their interactions are excluded from this table, even though they are more explanatory than the isotope end members.




